The benefits of reporting critical effect size values

### 2024-12-11 paper/paperDL_old.Rmd
FG reply to DL and adding new files (old are _old)

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample and has been targeted in the series of replications by the Open Science Collaboration [@open2015estimating], while the second study had a very large sample, and has attracted attention of the scientific community. The first case pertains to "Study 5" conducted by @mccrea2008self. This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, $t(26) = 1.87$, $p$ (one-tailed) $< .05$. Based on the reported data, Cohen's $d = 0.736$, the findings indicated a considerable effect size. However, the critical value for Cohen's $d$ in this context was $0.645$, suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. As the study was not a Registered Report, selective reporting of statistically significant findings was common at the time of publication, and most effects in psychology are substantially smaller, the effect size estimate is likely upwardly biased. Subsequent replication efforts [@open2015estimating] reaffirmed the significance of the effect, and used a somewhat larger sample ($n = 61$), $t(59)=2.325$. The estimated effect size was somehow lower than that originally reported, however, $d = 0.605$ (here, critical d = 0.428). THIS IS NOT A GOOD EXAMPLE, BECAUSE THESE TWO EFFECT SIZE ESTIMATES DO NOT DIFFER STATISTICALLY FROM EACH OTHER, SO WE CAN NOT SAY THE TWO EFFECT SIZES DIFFER! THERE SHOULD BE OTHER EXAMPLES IN THE LITERATURE, MAYBE FROM REGISTERED REPLICATION REPORTS.

The second scenario involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of $n = 689.003$, the researchers observed a statistically significant increase in the number of negative words typed by participants when positive posts were reduced in their timeline, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results from the study, the authors conclude that emotions expressed by other users on Facebook influence our own emotions [CHANGE A BIT, STAY CLOSER TO WHAT THE RESEARCHERS SAY, MAYBE QUOTE THEM DIRECTLY]. The critical values for Cohen's $d$ in this experimental design are $-0.006$ and $+0.006$. It should have been clear from the outset that the researchers should have considered the effect size beyond merely testing for the statistical significancee. The critical effect size should prompt reflection on which effect sizes are truly meaningful in this context, and whether the observed effect size ($d = 0.02$) is theoretically or practically interesting.

>>>>

\textcolor{red}{To provide a more tangible illustration of the practical applications of the critical effect sizes, let us examine two real-world instances drawn from published research. The first study was conducted on musicians and non musicians to detect differences in working memory, a modest sample size of 57 participants (42 musicians and 15 controls) was collected (@weiss2014spatial). The two groups were compared on a variety of tasks, related to verbal working memory (VWM) and tasks related to auditory skills. Regarding the VWM, the syllable-span task was administered and scores were computed for the maximal span and for the total number of sequences they correctly repeated. A two-tailed $t$-test was conducted to compare the groups. The resulting effect was statistically significant for the maximal span, $t(55) = -2.5$, $p = .017$, the Cohen's d calculated on summary statistics is of $-.74$, critical $d = \pm .62$. But for the total number of sequences the test did not reach statistical significance ($t(55) = -1.8$, $p = 0.076$), actual $d = -0.31$ and critical $d$ again $\pm.62$. Thus, given the limited sample size, a result must be of rather large magnitude to be associated with $p < 0.05$, while medium to small effect sizes will simply fail to reach significance. This case makes it obvious that a failure to find a significant result does not imply absence of an effect, while a statistically significant effect may represent an overestimation. }

The second scenario involves the study by @kramer2014experimental, which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of $n = 689.003$, the researchers observed a statistically significant increase in the number of negative words typed by participants when positive posts were reduced in their timeline, $t(310,044) = −5.63$, $p < 0.001$ (non-directional test). Based on this and other results from the study, the authors conclude that \textcolor{red}{"the emotions expressed by friends, via online social networks, influence our own moods, constituting, to our knowledge, the first experimental evidence for massive-scale emotional contagion via social networks".} The critical values for Cohen's $d$ in this experimental design are $-0.006$ and $+0.006$. It should have been clear from the outset that the researchers should have considered the effect size beyond merely testing for the statistical significance. The critical effect size should prompt reflection on which effect sizes are truly meaningful in this context, and whether the observed effect size ($d = 0.02$) is theoretically or practically interesting.

### 2024-12-11 paper/paperDL_old.tex
FG reply to DL and adding new files (old are _old)

To provide more tangible illustrations of the practical applications of critical effect size values, let us examine two real-world instances drawn from published research. The first one features a small sample and has been targeted in the series of replications by the Open Science Collaboration (Collaboration, 2015), while the second study had a very large sample, and has attracted attention of the scientific community. The first case pertains to ``Study 5'' conducted by McCrea (2008). This experiment compared two groups based on their performance percentage in a math test following a preliminary practice session with feedback of failure. One group was exposed to failure-excusing, self-handicapping thoughts, while both groups underwent the math test afterward. With a modest sample size of 28 participants split into two groups (13 and 15 participants, respectively), a directional one-tailed t-test was conducted on the percentage of correct answers to compare the groups. The resulting effect was statistically significant, \(t(26) = 1.87\), \(p\) (one-tailed) \(< .05\). Based on the reported data, Cohen's \(d = 0.736\), the findings indicated a considerable effect size. However, the critical value for Cohen's \(d\) in this context was \(0.645\), suggesting that, given this sample size, statistical significance is achieved only for estimated effects of substantial magnitude. As the study was not a Registered Report, selective reporting of statistically significant findings was common at the time of publication, and most effects in psychology are substantially smaller, the effect size estimate is likely upwardly biased. Subsequent replication efforts (Collaboration, 2015) reaffirmed the significance of the effect, and used a somewhat larger sample (\(n = 61\)), \(t(59)=2.325\). The estimated effect size was somehow lower than that originally reported, however, \(d = 0.605\) (here, critical d = 0.428). THIS IS NOT A GOOD EXAMPLE, BECAUSE THESE TWO EFFECT SIZE ESTIMATES DO NOT DIFFER STATISTICALLY FROM EACH OTHER, SO WE CAN NOT SAY THE TWO EFFECT SIZES DIFFER! THERE SHOULD BE OTHER EXAMPLES IN THE LITERATURE, MAYBE FROM REGISTERED REPLICATION REPORTS.

The second scenario involves the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689.003\), the researchers observed a statistically significant increase in the number of negative words typed by participants when positive posts were reduced in their timeline, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results from the study, the authors conclude that emotions expressed by other users on Facebook influence our own emotions {[}CHANGE A BIT, STAY CLOSER TO WHAT THE RESEARCHERS SAY, MAYBE QUOTE THEM DIRECTLY{]}. The critical values for Cohen's \(d\) in this experimental design are \(-0.006\) and \(+0.006\). It should have been clear from the outset that the researchers should have considered the effect size beyond merely testing for the statistical significancee. The critical effect size should prompt reflection on which effect sizes are truly meaningful in this context, and whether the observed effect size (\(d = 0.02\)) is theoretically or practically interesting.

@@

\leavevmode\vadjust pre{\hypertarget{ref-open2015estimating}{}}%
Collaboration, O. S. (2015). Estimating the reproducibility of psychological science. \emph{Science}, \emph{349}(6251), aac4716.

@@

\leavevmode\vadjust pre{\hypertarget{ref-mccrea2008self}{}}%
McCrea, S. M. (2008). Self-handicapping, excuse making, and counterfactual thinking: Consequences for self-esteem and future motivation. \emph{Journal of Personality and Social Psychology}, \emph{95}(2), 274.

>>>>

\textcolor{red}{To provide a more tangible illustration of the practical applications of the critical effect sizes, let us examine two real-world instances drawn from published research. The first study was conducted on musicians and non musicians to detect differences in working memory, a modest sample size of 57 participants (42 musicians and 15 controls) was collected (@weiss2014spatial). The two groups were compared on a variety of tasks, related to verbal working memory (VWM) and tasks related to auditory skills. Regarding the VWM, the syllable-span task was administered and scores were computed for the maximal span and for the total number of sequences they correctly repeated. A two-tailed $t$-test was conducted to compare the groups. The resulting effect was statistically significant for the maximal span, $t(55) = -2.5$, $p = .017$, the Cohen's d calculated on summary statistics is of $-.74$, critical $d = \pm .62$. But for the total number of sequences the test did not reach statistical significance ($t(55) = -1.8$, $p = 0.076$), actual $d = -0.31$ and critical $d$ again $\pm.62$. Thus, given the limited sample size, a result must be of rather large magnitude to be associated with $p < 0.05$, while medium to small effect sizes will simply fail to reach significance. This case makes it obvious that a failure to find a significant result does not imply absence of an effect, while a statistically significant effect may represent an overestimation. }

The second scenario involves the study by Kramer et al. (2014), which explored the impact of emotional content on Facebook users' experiences. With a notably large sample size of \(n = 689.003\), the researchers observed a statistically significant increase in the number of negative words typed by participants when positive posts were reduced in their timeline, \(t(310,044) = −5.63\), \(p < 0.001\) (non-directional test). Based on this and other results from the study, the authors conclude that \textcolor{red}{"the emotions expressed by friends, via online social networks, influence our own moods, constituting, to our knowledge, the first experimental evidence for massive-scale emotional contagion via social networks".} The critical values for Cohen's \(d\) in this experimental design are \(-0.006\) and \(+0.006\). It should have been clear from the outset that the researchers should have considered the effect size beyond merely testing for the statistical significance. The critical effect size should prompt reflection on which effect sizes are truly meaningful in this context, and whether the observed effect size (\(d = 0.02\)) is theoretically or practically interesting.

@@

@@
